\pkg{cubble}: An R Package for Structuring Spatio-temporal Data

### 2022-04-18 1-intro.Rmd
a new skeleton of intro incorporating various past writing

Spatio-temporal data involves variables observed in the space across time and this paper focuses on spatio-temporal vector data with fixed locations and regular time intervals. This type of data can naturally be considered as a data cube with three axes: spatial identifier, temporal identifier, and variables. In this cubic formation, space only occupies one axis, rather than two, to avoid hypercubes for multivariate spatio-temporal data. While conceptually the data format is relatively simple, the data objects coming to the analysts can vary in forms in the analysis. Figure \ref{fig:illu-input} shows three examples of wild spatio-temporal data: 1) a single table that includes both the spatial and temporal variables; 2) separate tables for spatial and temporal variables; 3) array-based data, i.e. NetCDF (Network Common Data Form) data, which is commonly used in earth science, e.g. climatology, meteorology, and oceanography, among others, to organise multivariate data in a spatio-temporal grid.

```{r illu-input, echo = FALSE, out.width = "100%", fig.cap="An illustration of different spatio-temporal data layouts: (1) a single table for both spatial and temporal variables; (2) separate tables for spatial and temporal variables; and (3) an array data structure that populates variables into a gridded longitude-latitude-time (hyper)cube.", fig.align='center'}
knitr::include_graphics(here::here("figures/diagram-keynotes/diagram-keynotes.001.png"))
```

@@

While incorporating both spatial and temporal information enriches the data in the analysis, spatio-temporal data can sometimes be hard to work with given its two observational units in the variables. The single table format in Figure \ref{fig:illu-input} is convenient for spatio-temporal operations that involve both types of variables, however, it makes duplicates of spatial variables and is not ideal when only looking at the spatial aspect of the data. The two tables format allows separate operations on spatial and temporal variables while analysts need to work with two data objects simultaneously and join them together into the single table format for spatio-temporal operations. Neither of the two formats is desirable at the same time for spatial, temporal, and spatio-temporal operations. This issue implies analysts need to constantly think about the data format and arrange variables into the proper shape before the actual operations, which can be a laborious step when switching between the two formats back and forth several times during the analysis.

@@

In the \proglang{R} community, many data objects have been proposed to work with specific spatio-temporal tasks, while few can freely explore the data. \pkg{spacetime} [@spacetime] is an attempt for such a data object, but the spatial and temporal classes it is built on have now been gradually replaced by more recent implementations. The \pkg{stars} [@stars] package uses an array to wrangle spatio-temporal data, but the dense array it uses can quickly exhaust the memory for data in the case of irregular spatial grid. While there has not yet been a data object that makes it easy to wrangle spatio-temporal data, there is a strong need for such tools given the ubiquitous spatial and temporal data collected in the society, i.e., climate observations from weather stations, air quality variables from air monitoring stations. The demand for such tools can also come from spatial data analysts wishing to incorporate temporal data in their analysis or vice versa.

This paper presents a new \proglang{R} package, \pkg{cubble}, to overcome the problem with organising variables in different levels when working with spatio-temporal data. In the package a new data structure, also called \code{cubble}, is proposed to organise spatial and temporal variables as two forms of a single data object so that they can be wrangled separately while synchronised as a whole piece. With this new data object, analysts can spend less time writing codes to organise spatial and temporal variables and focus more on the exploratory data analysis itself. The software is available from the Comprehensive R Archive Network (CRAN) at [CRAN link].

The rest of the paper is divided as follows: Section 2 presents the main design and functionality of \pkg{cubble}. Section 3 explains how cubble deals with more advanced considerations including data with hierarchical structure, data matching, how cubble fits with existing static and interactive visualisation tools, and spatio-temporal data transformation. Section 4 uses Australia weather station data and river level data as examples to demonstrate the use of \pkg{cubble}. An example on how \pkg{cubble} handles NetCDF data is also provided. Section 5 concludes the paper.

>>>>

Spatio-temporal data has a spatial component referring to the location of each observation, a temporal component that is recorded at regular or irregular time intervals, and multiple variables measured at each spatial and temporal value. In data collection, spatio-temporal data is less often collected as a whole piece but fragmented spatial and temporal data for analysts to ensemble into spatio-temporal data. With the [integrated] data, we hope to fix the time to explore the spatial distribution, or fix the spaital location to explore temporal trends. Ideally, we would be able to carry out each of these tasks easily, and even accommodate complex explorations encompassing both simultaneously. These are all underpinned by a [flexible] data structure that can represent spatio-temporal data.

The *Representing data* section of the CRAN task view [ref] lists existing spatio-temporal data representation. Among them, @spacetime proposes four layouts: full grid layout, sparse grid layout, irregular layout, and trajectory. These layouts are implemented in the \pkg{spacetime} package with underlying class \pkg{sp} [@sp] and \pkg{xts} [@xts] for handling spatial and temporal variables. The \pkg{stars} package has been a recent implementation to handle spatio-temporal data in an array. The package depends on the \pkg{sf} package and has allowed more access to tidyverse verbs in the workflow.

What remains to be difficult when wrangling spatio-temporal data is that the raw data is not naturally structured in any one of the representation mentioned above and the first difficulty is to cast the data, potentially from multiple sources into one of the data representation. Some examples of these includes map data recorded the shape of a collection of area of interest is provided by the government website; geo-scientific data recorded with longitude and latitude coordinate of locations in the area; and temporal variables indexed by both location and time for the temporal aspect. For spatio-temporal data analysis, merging all these sources can be challenging and joining these tables can go wrong because of the slightest unmatched of the linking variable from different sources.

Another issue is how the concept of tidy data should be applied to spatio-temporal data. The tidy data concept advocates the data to be shaped into 1) one row per observation; 2) one column per variable; and 3) one type of data per table. The long form data is preferred over wide data given the downstream data cleaning and visualisation software i.e. \pkg{dplyr} and \pkg{ggplot2}. However, the long form can be inefficient to store feature geometries, especially for large multipolygon over long time period, i.e. sub-daily time series data.

This paper presents a new \proglang{R} package, \pkg{cubble}, to [overcome the problem] with organising variables in different levels when working with spatio-temporal data. The data structure focuses on connecting to the upstream data collection and is suitable for temporal data collected at the fixed location over time. Among the four spacetime layouts in @spacetime, it can be full grid layout, sparse grid layout, or irregular layout, but trajectory is not within the scope of this work. In the package a new data structure, also called \code{cubble}, is proposed to organise spatial and temporal variables as two forms of a single data object so that they can be wrangled separately while synchronised as a whole piece. With this new data object, analysts can spend less time writing codes to organise spatial and temporal variables and focus more on the exploratory data analysis itself. The software is available from the Comprehensive R Archive Network (CRAN) at [CRAN link].

The rest of the paper is divided as follows: Section \@ref(cube) discusses the spatio-temporal cube to conceptualise spatio-temporal data. Section \@ref(cubble) presents the main design and functionality of \pkg{cubble}. Section \@ref(others) explains how cubble deals with more advanced considerations including data with hierarchical structure, data matching, how cubble fits with existing static and interactive visualisation tools, and spatio-temporal data transformation. Section \@ref(examples) uses Australia weather station data and river level data as examples to demonstrate the use of \pkg{cubble}. An example on how \pkg{cubble} handles NetCDF data is also provided. Section \@ref(conclude) concludes the paper.

<!-- there is a strong need for such tools given the ubiquitous spatial and temporal data collected in the society, i.e., climate observations from weather stations, air quality variables from air monitoring stations. The demand for such tools can also come from spatial data analysts wishing to incorporate temporal data in their analysis or vice versa. -->

<!-- This type of data can naturally be considered as a data cube with three axes: spatial identifier, temporal identifier, and variables. In this cubic formation, space only occupies one axis, rather than two, to avoid hypercubes for multivariate spatio-temporal data. While conceptually the data format is relatively simple, the data objects coming to the analysts can vary in forms in the analysis. Figure \ref{fig:illu-input} shows three examples of wild spatio-temporal data: 1) a single table that includes both the spatial and temporal variables; 2) separate tables for spatial and temporal variables; 3) array-based data, i.e. NetCDF (Network Common Data Form) data, which is commonly used in earth science, e.g. climatology, meteorology, and oceanography, among others, to organise multivariate data in a spatio-temporal grid. -->

```{r illu-input, echo = FALSE, out.width = "100%", fig.cap="An illustration of different spatio-temporal data layouts: (1) a single table for both spatial and temporal variables; (2) separate tables for spatial and temporal variables; and (3) an array data structure that populates variables into a gridded longitude-latitude-time (hyper)cube.", fig.align='center'}
#knitr::include_graphics(here::here("figures/diagram-keynotes/diagram-keynotes.001.png"))
```

@@

<!-- While incorporating both spatial and temporal information enriches the data in the analysis, spatio-temporal data can sometimes be hard to work with given its two observational units in the variables. The single table format in Figure \ref{fig:illu-input} is convenient for spatio-temporal operations that involve both types of variables, however, it makes duplicates of spatial variables and is not ideal when only looking at the spatial aspect of the data. The two tables format allows separate operations on spatial and temporal variables while analysts need to work with two data objects simultaneously and join them together into the single table format for spatio-temporal operations. Neither of the two formats is desirable at the same time for spatial, temporal, and spatio-temporal operations. This issue implies analysts need to constantly think about the data format and arrange variables into the proper shape before the actual operations, which can be a laborious step when switching between the two formats back and forth several times during the analysis. -->

@@

<!-- The package \pkg{sf} [@sf] uses data frame objects to represent spatial data with an additional \code{geometry} column dedicating to feature geometries [@jr_herring_opengis_2011] (e.g. POINT, MULTIPOINT, POLYGON, MULTIPOLYGON etc).  -->

<!-- For temporal data, \pkg{tsibble} [@tsibble] builds on top of the \pkg{tibble} [@tibble] object and explicitly includes date and time in columns as variables, rather than data attributes.  -->

<!-- Underlying tidyverse is the concept of tidy data [@tidydata], which considers how variables should be structured for analysis. In the three formats (time-wide, space-wide, and long) summarised in the @spacetime paper, only the long form is considered as tidy. The main advantages of it over the time-wide and space-wide format is that the latter treat either time or location as variable names (Section 3.1 in @tidydata) rather than values of variables, making them difficult to be wrangled. One problem with the long format is that it can be inefficient to store feature geometries, especially for large MULTIPOLYGON over long time period.  -->

<!-- For spatial data, \pkg{sf} [@sf] covers a wide range of geometrical operations and is the format most map data stored. When incorporating the time dimension, a time wide format, introduced in @spacetime, is usually adopted. This treats time as variable name rather than value and makes it inconvenient to wrangle time, as per the tidy data principle [@tidydata]. In the temporal domain, \pkg{tsibble} [@tsibble] follows the tidy data principle to arrange the time in the long format, while when adding spatial variables, a long format implies repeating variables at each time and this can make a huge difference when the repetition is on large geometrical shapes with long time period.  -->

<!-- In the \proglang{R} community, many data objects have been proposed to work with specific spatio-temporal tasks, while few can freely explore the data. \pkg{spacetime} [@spacetime] is an attempt for such a data object, but the spatial and temporal classes it is built on have now been gradually replaced by more recent implementations. The \pkg{stars} [@stars] package uses an array to wrangle spatio-temporal data, but the dense array it uses can quickly exhaust the memory for data in the case of irregular spatial grid. -->

### 2022-04-18 2-0-cube.Rmd
a new skeleton of intro incorporating various past writing

# Conceptual framework

Multivariate spatio-temporal data can be conceptualised using a cubical data model with reference to variables, time and space. This naturally motivates the usage of array to represent and store spatio-temporal data, as evident by satellite imageries, large climate models, among others. The work by @lu_multidimensional_2018 has discussed the suitability of this representation through array algebra and practical spatio-temporal tasks and software implementation has been seen in many R packages: \pkg{raster} [@raster], \pkg{terra} [@terra], and \pkg{stars} [@stars].

While it is efficient to represent spatio-temporal data in an array, there are situation where array [is not the best solution]. mutate in the array is not clearly define (how to add a column vector)

Also, array is not the best solution for a large subset of data, specifically, the vector spatio-temporal data. With vector data, spatial and temporal variables are commonly collected in separate 2D tables and there is usually a location ID to link the spatial and temporal information. Sometimes, the geometrical shape of the location may not present and an additional linking is needed to incorporate the map data. With the multiple pieces in 2D data frame format, there should be a solution to organise all these data from multiple sources in a 2D structure, without needing to convert it into higher dimensional representation.

Reliable software and data structure has been proposed in the spatial and temporal domain, while a direct extension to spatio-temporal data is not ideal. The main issue here lays in the format that spatial and temporal data is stored. For spatial data, \pkg{sf} [@sf] covers a wide range of geometrical operations and is the format most map data stored. When incorporating the time dimension, a time wide format, introduced in @spacetime, is usually adopted. This treats time as variable name rather than value and makes it inconvenient to wrangle time, as per the tidy data principle [@tidydata]. In the temporal domain, \pkg{tsibble} [@tsibble] follows the tidy data principle to arrange the time in the long format, while when adding spatial variables, a long format implies repeating variables at each time and this can make a huge difference when the repetition is on large geometrical shapes with long time period.

>>>>

<!-- and software implementation has been seen in many R packages: \pkg{raster} [@raster], \pkg{terra} [@terra], and \pkg{stars} [@stars]. -->

<!-- # Existing work and new challenges -->

<!-- Spatio-temporal software should echo with the advance in spatial and temporal data analysis. This section will start with the \pkg{spacetime} paper [@spacetime] that sets the foundation for wrangling spatio-temporal data in \proglang{R}, and then describe some emerging concepts and software in data analysis since the paper publishes and what remains as challenges. -->

<!-- Also, array is not the best solution for a large subset of data, specifically, the vector spatio-temporal data. With vector data, spatial and temporal variables are commonly collected in separate 2D tables and there is usually a location ID to link the spatial and temporal information. Sometimes, the geometrical shape of the location may not present and an additional linking is needed to incorporate the map data. With the multiple pieces in 2D data frame format, there should be a solution to organise all these data from multiple sources in a 2D structure, without needing to convert it into higher dimensional representation. -->

# Conceptual framework: spatio-temporal cube {#cube}

Multivariate spatio-temporal data can be conceptualised using a cubical data model with reference to variables, time and space. This naturally motivates the usage of array to represent and store spatio-temporal data, as evident by satellite imageries, large climate models, among others. The work by @lu_multidimensional_2018 has discussed the suitability of this representation through array algebra and practical spatio-temporal tasks.

The cubical conceptual framework provides a generalisation to the operation can be done on the data for visualisation. A paper by @bach_review_2014 has reviewed the temporal data visualisation based on space-time cube operations. Notice that the term space-time cube in their article "does not need to involve spatial data", but referring to "an abstract 2D substrate that is used to visualize data at a specific time". Despite its main focus is on temporal data, the mindset of abstracting out data representation to construct visualisation is applicable to our spatio-temporal data manipulation and visualisation.

### 2022-04-18 5-conclusion.Rmd
clean up bibbib
minor al layout adjustment
a new skeleton of intro incorporating various past writing

---
title: "5-conclusion"
output: pdf_document
---
# Conclusion

@@

Further development and maintenance of the package involves responding to changes in the tidyverse packages that \pkg{cubble} imports, in particular, \pkg{tibble}, \pkg{tidyr}, and \pkg{dplyr}. Another area for further development is to extend \pkg{cubble} to other spatial objects other than points.

@@

>>>>

---
title: "5-conclusion"
output: pdf_document
---
# Conclusion {#conclude}

@@

Further development and maintenance of the package involves responding to changes in the tidyverse packages that \pkg{cubble} imports, in particular, \pkg{tibble}, \pkg{tidyr}, and \pkg{dplyr}. In the spatial aspect, the simple feature representation wraps spatial coordinates in a list column, for some complex spatial operations, while sometimes, analysts may want to unpack them into longitude and latitude columns to work from. Another extension to \pkg{cubble} is to build a smooth transition between the coordinate columns and simple feature geometry column.

\newpage

@@

\newpage

### 2022-04-18 jss/cubble-jss.tex
a new skeleton of intro incorporating various past writing

@@

Spatio-temporal data involves variables observed in the space across
time and this paper focuses on spatio-temporal vector data with fixed
locations and regular time intervals. This type of data can naturally be
considered as a data cube with three axes: spatial identifier, temporal
identifier, and variables. In this cubic formation, space only occupies
one axis, rather than two, to avoid hypercubes for multivariate
spatio-temporal data. While conceptually the data format is relatively
simple, the data objects coming to the analysts can vary in forms in the
analysis. Figure \ref{fig:illu-input} shows three examples of wild
spatio-temporal data: 1) a single table that includes both the spatial
and temporal variables; 2) separate tables for spatial and temporal
variables; 3) array-based data, i.e.~NetCDF (Network Common Data Form)
data, which is commonly used in earth science, e.g.~climatology,
meteorology, and oceanography, among others, to organise multivariate
data in a spatio-temporal grid.

{\centering \includegraphics[width=1\linewidth]{/Users/sherryzhang/Documents/research/paper-cubble/figures/diagram-keynotes/diagram-keynotes.001}

\caption[An illustration of different spatio-temporal data layouts]{An illustration of different spatio-temporal data layouts: (1) a single table for both spatial and temporal variables; (2) separate tables for spatial and temporal variables; and (3) an array data structure that populates variables into a gridded longitude-latitude-time (hyper)cube.}\label{fig:illu-input}
\end{figure}
\end{CodeChunk}

While incorporating both spatial and temporal information enriches the
data in the analysis, spatio-temporal data can sometimes be hard to work
with given its two observational units in the variables. The single
table format in Figure \ref{fig:illu-input} is convenient for
spatio-temporal operations that involve both types of variables,
however, it makes duplicates of spatial variables and is not ideal when
only looking at the spatial aspect of the data. The two tables format
allows separate operations on spatial and temporal variables while
analysts need to work with two data objects simultaneously and join them
together into the single table format for spatio-temporal operations.
Neither of the two formats is desirable at the same time for spatial,
temporal, and spatio-temporal operations. This issue implies analysts
need to constantly think about the data format and arrange variables
into the proper shape before the actual operations, which can be a
laborious step when switching between the two formats back and forth
several times during the analysis.

In the \proglang{R} community, many data objects have been proposed to
work with specific spatio-temporal tasks, while few can freely explore
the data. \pkg{spacetime} \citep{spacetime} is an attempt for such a
data object, but the spatial and temporal classes it is built on have
now been gradually replaced by more recent implementations. The
\pkg{stars} \citep{stars} package uses an array to wrangle
spatio-temporal data, but the dense array it uses can quickly exhaust
the memory for data in the case of irregular spatial grid. While there
has not yet been a data object that makes it easy to wrangle
spatio-temporal data, there is a strong need for such tools given the
ubiquitous spatial and temporal data collected in the society, i.e.,
climate observations from weather stations, air quality variables from
air monitoring stations. The demand for such tools can also come from
spatial data analysts wishing to incorporate temporal data in their
analysis or vice versa.

This paper presents a new \proglang{R} package, \pkg{cubble}, to
overcome the problem with organising variables in different levels when
working with spatio-temporal data. In the package a new data structure,
also called \code{cubble}, is proposed to organise spatial and temporal
variables as two forms of a single data object so that they can be
wrangled separately while synchronised as a whole piece. With this new
data object, analysts can spend less time writing codes to organise
spatial and temporal variables and focus more on the exploratory data
analysis itself. The software is available from the Comprehensive R
Archive Network (CRAN) at {[}CRAN link{]}.

The rest of the paper is divided as follows: Section 2 presents the main
design and functionality of \pkg{cubble}. Section 3 explains how cubble
deals with more advanced considerations including data with hierarchical
structure, data matching, how cubble fits with existing static and
interactive visualisation tools, and spatio-temporal data
transformation. Section 4 uses Australia weather station data and river
level data as examples to demonstrate the use of \pkg{cubble}. An
example on how \pkg{cubble} handles NetCDF data is also provided.
Section 5 concludes the paper.

\hypertarget{conceptual-framework-spatio-temporal-cube}{%
\section{Conceptual framework: spatio-temporal
cube}\label{conceptual-framework-spatio-temporal-cube}}

@@

\hypertarget{existing-work-and-new-challenges}{%
\section{Existing work and new
challenges}\label{existing-work-and-new-challenges}}

To represent spatio-temporal data, \citet{spacetime} proposes four
layouts: full grid layout (STF), sparse grid layout (STS), irregular
layout (STI), and trajectory (STT). These layouts are implemented in the
\pkg{spacetime} package with underlying class \pkg{sp} \citep{sp} and
\pkg{xts} \citep{xts} for handling spatial and temporal variables. While
these layouts and classes are still being used and well-maintained
today, they are not compatible with the \pkg{tidyverse} ecosystem
\citep{tidyverse}, which has a huge influence on data analysis in
\proglang{R}, including in the spatial and temporal communities. The
package \pkg{sf} \citep{sf} uses data frame objects to represent spatial
data with an additional \code{geometry} column dedicating to feature
geometries \citep{jr_herring_opengis_2011} (e.g.~POINT, MULTIPOINT,
POLYGON, MULTIPOLYGON etc). For temporal data, \pkg{tsibble}
\citep{tsibble} builds on top of the \pkg{tibble} \citep{tibble} object
and explicitly includes date and time in columns as variables, rather
than data attributes.

Underlying tidyverse is the concept of tidy data \citep{tidydata}, which
considers how variables should be structured for analysis. In the three
formats (time-wide, space-wide, and long) summarised in the
\citet{spacetime} paper, only the long form is considered as tidy. The
main advantages of it over the time-wide and space-wide format is that
the latter treat either time or location as variable names (Section 3.1
in \citet{tidydata}) rather than values of variables, making them
difficult to be wrangled. One problem with the long format is that it
can be inefficient to store feature geometries, especially for large
MULTIPOLYGON over long time period. This motivates a new tidyverse
compatible data object to work with spatio-temporal data.

\textbf{two challenages: multiple data sources + large spatio-temporal
object} In additional to the shift into more tidyverse workflow for
working with data, spatio-temporal data have some unique challenges. The
first one being the split of variables in multiple tables. Often the
time, spatio-temporal data is not presented in a single data object, but
split into fragmented pieces from different sources. For example, map
data recorded the shape of a collection of area of interest is provided
by the government website; the geo-scientific data is recorded with
longitude and latitude coordinate of locations in the area; and temporal
variables indexed by both location and time need to be further queried
using location data. For spatio-temporal data analysis, merging all
these sources can be challenging and joining these tables can go wrong
because of the slightest unmatched of the linking variable from
different sources.

Another challenge when analysing spatio-temporal data is on the simple
feature representation of spatial data. While the simple feature is able
to handle large and complex spatial geometry and perform precise spatial
operations, data analysts usually do not have a large expectation on the
map accuracy and would prefer a simplified map with quick rendering for
visualisation. Also, the feature geometry column in \pkg{sf} wraps the
spatial coordinates in a list column, sometimes, analysts may like to
unpack them into longitude and latitude to work with. A smooth
transition with the coordinate columns and and simple feature geometry
column would satisfy analysts from both ends.

With some of the aforementioned issues recognised, the \pkg{stars}
package has been a new implementation to handle spatio-temporal data
conceptually as data cubes and implemented as arrays. The package
depends on the \pkg{sf} package and has allowed more access to tidyverse
verbs in the workflow. However, for vector spatio-temporal data, major
difficulties to work with \pkg{stars} is first to cast the variables
from multiple tables into a star object and then keep the operations in
the high-dimensional cube. Given that most table mentioned before
collect variables in a 2D table, analysts could benefit from a structure
to allow them work in 2D tables.

\hypertarget{the-cubble-package}{%
\section{The cubble package}\label{the-cubble-package}}

@@

\hypertarget{other-features-and-considerations}{%
\section{Other features and
considerations}\label{other-features-and-considerations}}

@@

\hypertarget{conclusion}{%
\section{Conclusion}\label{conclusion}}

@@

Further development and maintenance of the package involves responding
to changes in the tidyverse packages that \pkg{cubble} imports, in
particular, \pkg{tibble}, \pkg{tidyr}, and \pkg{dplyr}. Another area for
further development is to extend \pkg{cubble} to other spatial objects
other than points.

@@

\bibliography{references.bib}

>>>>

% From pandoc table feature
\usepackage{longtable,booktabs,array}
\usepackage{calc} % for calculating minipage widths
% Correct order of tables after \paragraph or \subparagraph
\usepackage{etoolbox}
\makeatletter
\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
\makeatother
% Allow footnotes in longtable head/foot
\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
\makesavenoteenv{longtable}

@@

Spatio-temporal data has a spatial component referring to the location of each observation, a temporal component that is recorded at regular or irregular time intervals, and multiple variables measured at each spatial and temporal value. In data collection, spatio-temporal data is less often collected as a whole piece but fragmented spatial and temporal data for analysts to ensemble into spatio-temporal data. With the {[}integrated{]} data, we hope to fix the time to explore the spatial distribution, or fix the spaital location to explore temporal trends. Ideally, we would be able to carry out each of these tasks easily, and even accommodate complex explorations encompassing both simultaneously. These are all underpinned by a {[}flexible{]} data structure that can represent spatio-temporal data.

The \emph{Representing data} section of the CRAN task view {[}ref{]} lists existing spatio-temporal data representation. Among them, \citet{spacetime} proposes four layouts: full grid layout, sparse grid layout, irregular layout, and trajectory. These layouts are implemented in the \pkg{spacetime} package with underlying class \pkg{sp} \citep{sp} and \pkg{xts} \citep{xts} for handling spatial and temporal variables. The \pkg{stars} package has been a recent implementation to handle spatio-temporal data in an array. The package depends on the \pkg{sf} package and has allowed more access to tidyverse verbs in the workflow.

What remains to be difficult when wrangling spatio-temporal data is that the raw data is not naturally structured in any one of the representation mentioned above and the first difficulty is to cast the data, potentially from multiple sources into one of the data representation. Some examples of these includes map data recorded the shape of a collection of area of interest is provided by the government website; geo-scientific data recorded with longitude and latitude coordinate of locations in the area; and temporal variables indexed by both location and time for the temporal aspect. For spatio-temporal data analysis, merging all these sources can be challenging and joining these tables can go wrong because of the slightest unmatched of the linking variable from different sources.

Another issue is how the concept of tidy data should be applied to spatio-temporal data. The tidy data concept advocates the data to be shaped into 1) one row per observation; 2) one column per variable; and 3) one type of data per table. The long form data is preferred over wide data given the downstream data cleaning and visualisation software i.e.~\pkg{dplyr} and \pkg{ggplot2}. However, the long form can be inefficient to store feature geometries, especially for large multipolygon over long time period, i.e.~sub-daily time series data.

This paper presents a new \proglang{R} package, \pkg{cubble}, to {[}overcome the problem{]} with organising variables in different levels when working with spatio-temporal data. The data structure focuses on connecting to the upstream data collection and is suitable for temporal data collected at the fixed location over time. Among the four spacetime layouts in \citet{spacetime}, it can be full grid layout, sparse grid layout, or irregular layout, but trajectory is not within the scope of this work. In the package a new data structure, also called \code{cubble}, is proposed to organise spatial and temporal variables as two forms of a single data object so that they can be wrangled separately while synchronised as a whole piece. With this new data object, analysts can spend less time writing codes to organise spatial and temporal variables and focus more on the exploratory data analysis itself. The software is available from the Comprehensive R Archive Network (CRAN) at {[}CRAN link{]}.

The rest of the paper is divided as follows: Section \ref{cube} discusses the spatio-temporal cube to conceptualise spatio-temporal data. Section \ref{cubble} presents the main design and functionality of \pkg{cubble}. Section \ref{others} explains how cubble deals with more advanced considerations including data with hierarchical structure, data matching, how cubble fits with existing static and interactive visualisation tools, and spatio-temporal data transformation. Section \ref{examples} uses Australia weather station data and river level data as examples to demonstrate the use of \pkg{cubble}. An example on how \pkg{cubble} handles NetCDF data is also provided. Section \ref{conclude} concludes the paper.

\hypertarget{cube}{%
\section{Conceptual framework: spatio-temporal cube}\label{cube}}

@@

\hypertarget{cubble}{%
\section{The cubble package}\label{cubble}}

@@

\hypertarget{others}{%
\section{Other features and considerations}\label{others}}

@@

\hypertarget{conclude}{%
\section{Conclusion}\label{conclude}}

@@

Further development and maintenance of the package involves responding to changes in the tidyverse packages that \pkg{cubble} imports, in particular, \pkg{tibble}, \pkg{tidyr}, and \pkg{dplyr}. In the spatial aspect, the simple feature representation wraps spatial coordinates in a list column, for some complex spatial operations, while sometimes, analysts may want to unpack them into longitude and latitude columns to work from. Another extension to \pkg{cubble} is to build a smooth transition between the coordinate columns and simple feature geometry column.

@@

\bibliography{../references.bib}
